{fairmetrics}: An R package for group fairness evaluation

### 2025-06-02 paper.Rmd
add space before citation

ML models are increasingly integrated into high-stakes domains to support decision making that significantly impacts individuals and society more broadly, including criminal justice, healthcare, finance, employment, and education[@mehrabi_survey_21].  Mounting evidence suggest that these models often exhibit bias across groups defined by protected attributes. For example, within criminal justice, the Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) software, a tool used by U.S. courts to evaluate the risk of defendants becoming recidivists, was found to incorrectly classify Black defendants as high-risk at nearly twice the rate of white defendants[@mattuMachineBias].  This bias impacted Black defendants by potentially leading to harsher bail decisions, longer sentences, and reduced parole opportunities compared to white defendants with similar risk profiles. Similarly, within healthcare, a commercial risk-prediction algorithm deployed in the U.S. to identify patients with complex health needs for high-risk care management programs was shown to be significantly less calibrated for Black patients relative to white patients[@obermeyerDissectingRacialBias2019]. This caused Black patients with equivalent health conditions to be under-referred for essential care services compared to white patients.  These examples illustrate that there is an urgent need for practitioners and researchers to ensure that ML models support fair decision making before they are deployed in real-world applications.

@@

The `{fairmetrics}` package is designed to evaluate group fairness in the setting of binary classification with a binary protected attribute. This restriction reflects standard practice in the fairness literature and is motivated by several considerations. First, binary classification remains prevalent in many high-stakes applications, such as loan approval, hiring decisions, and disease screening, where outcomes are typically framed as accept/reject or positive/negative[@mehrabi_survey_21]. Second, group fairness is the most widely used framework for binary classification tasks[@mehrabi_survey_21]. Third, when protected attributes have more than two categories, there is no clear consensus on how to evaluate group fairness[NOTE:citation to add: https://arxiv.org/abs/2205.05770]. This focus enables {fairmetrics} to provide statistically grounded uncertainty quantification for group fairness metrics commonly applied in binary classification tasks across diverse application domains.

@@

Group fairness criteria are primarily classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022][NOTE: cite https://arxiv.org/abs/2406.09307]. Independence requires that the model's classifications be statistically independent of the protected attribute, meaning the likelihood of receiving a positive prediction is the same across protected groups. Separation requires independence between the classifications and the protected attribute conditional on the true outcome, so that the probability of a positive prediction is equal across protected groups within the positive (or negative) outcome class. Sufficiency requires independence between the outcome and the protected attribute conditional on the prediction, implying that once the model's prediction is known, the protected attribute provides no additional information about the true outcome.  Below we summarize the fairness metrics that are available within the {fairmetrics} package.

@@

Other R packages similar to {fairmetrics} include {fairness}[@fairness_package], {fairmodels} [@wisniewski2022fairmodels] and {mlr3fairness}[mlr3fairness_package]. The differences between {fairmetrics} and these other packages is twofold. The primary difference between is that {fairmetrics} calculates the ratio and difference between group fairness criterion and their corresponding confidence intervals of fairness metrics via bootstrap, allowing for more meaningful inferences about the fairness criteria. Additionally, in contrast to the {fairmodels}, {fairness} and {mlr3fairness} packages, the {fairmetrics} package does not posses any external dependencies and has a lower memory footprint, resulting in an environment agnostic tool that can be used with modest hardware and older systems. \hyperref[tab:memory_dep_usage]{Table~\ref*{tab:memory_dep_usage}} shows the comparison of memory used and dependencies required when loading each library.

>>>>

ML models are increasingly integrated into high-stakes domains to support decision making that significantly impacts individuals and society more broadly, including criminal justice, healthcare, finance, employment, and education [@mehrabi_survey_21].  Mounting evidence suggest that these models often exhibit bias across groups defined by protected attributes. For example, within criminal justice, the Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) software, a tool used by U.S. courts to evaluate the risk of defendants becoming recidivists, was found to incorrectly classify Black defendants as high-risk at nearly twice the rate of white defendants [@mattuMachineBias].  This bias impacted Black defendants by potentially leading to harsher bail decisions, longer sentences, and reduced parole opportunities compared to white defendants with similar risk profiles. Similarly, within healthcare, a commercial risk-prediction algorithm deployed in the U.S. to identify patients with complex health needs for high-risk care management programs was shown to be significantly less calibrated for Black patients relative to white patients [@obermeyerDissectingRacialBias2019]. This caused Black patients with equivalent health conditions to be under-referred for essential care services compared to white patients.  These examples illustrate that there is an urgent need for practitioners and researchers to ensure that ML models support fair decision making before they are deployed in real-world applications.

@@

The `{fairmetrics}` package is designed to evaluate group fairness in the setting of binary classification with a binary protected attribute. This restriction reflects standard practice in the fairness literature and is motivated by several considerations. First, binary classification remains prevalent in many high-stakes applications, such as loan approval, hiring decisions, and disease screening, where outcomes are typically framed as accept/reject or positive/negative [@mehrabi_survey_21]. Second, group fairness is the most widely used framework for binary classification tasks [@mehrabi_survey_21]. Third, when protected attributes have more than two categories, there is no clear consensus on how to evaluate group fairness [@lum_bias_22]. This focus enables {fairmetrics} to provide statistically grounded uncertainty quantification for group fairness metrics commonly applied in binary classification tasks across diverse application domains.

@@

Group fairness criteria are primarily classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022, @Gao_Chou_McCaw_Thurston_Varghese_Hong_Gronsbell_2024]. Independence requires that the model's classifications be statistically independent of the protected attribute, meaning the likelihood of receiving a positive prediction is the same across protected groups. Separation requires independence between the classifications and the protected attribute conditional on the true outcome, so that the probability of a positive prediction is equal across protected groups within the positive (or negative) outcome class. Sufficiency requires independence between the outcome and the protected attribute conditional on the prediction, implying that once the model's prediction is known, the protected attribute provides no additional information about the true outcome.  Below we summarize the fairness metrics that are available within the {fairmetrics} package.

@@

Other R packages similar to {fairmetrics} include {fairness} [@fairness_package], {fairmodels} [@wisniewski2022fairmodels] and {mlr3fairness} [@mlr3fairness_package]. The differences between {fairmetrics} and these other packages is twofold. The primary difference between is that {fairmetrics} calculates the ratio and difference between group fairness criterion and their corresponding confidence intervals of fairness metrics via bootstrap, allowing for more meaningful inferences about the fairness criteria. Additionally, in contrast to the {fairmodels}, {fairness} and {mlr3fairness} packages, the {fairmetrics} package does not posses any external dependencies and has a lower memory footprint, resulting in an environment agnostic tool that can be used with modest hardware and older systems. \hyperref[tab:memory_dep_usage]{Table~\ref*{tab:memory_dep_usage}} shows the comparison of memory used and dependencies required when loading each library.
